# Analysis of longitudinal data


```{r}
date()
```


```{r}
# Loading packages
library(dplyr)
library(tidyr)
library(ggplot2)
```


## Part 1. RATS data


```{r}
# Reading the data set
RATSL <- read.csv('data/RATSL.csv')

# Looking at the structure of the data set
str(RATSL)

# Converting variables as factors
RATSL$ID <- factor(RATSL$ID)
RATSL$Group <-factor(RATSL$Group)
RATSL$WD <-factor(RATSL$WD)

str(RATSL)
```
Let's look at the plots of the data.

```{r}
# Draw the plot:
ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))
```

We can see that group 1 has the smallest average weight. Group two and group three have a similar weight. There's clearly visible outlier in the second group which has much higher weight then the rest of the rats in that group. In time the weight of the rats grows.




```{r}
# Standartizing rats weight
RATSL <- RATSL %>% group_by(Time) %>%
  mutate(stdweight = (Weight - mean(Weight))/sd(Weight) ) %>% ungroup()

# Looking at the standartized data
glimpse(RATSL)

# Plotting again 

ggplot(RATSL, aes(x = Time, y = stdweight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "standardized weight")

```

After standardazing the weight, the growing trend is not anymore so visible in the data.



```{r}
# Setting the number of times
n <- RATSL$Time %>% unique() %>% length()

# Summarizing the data with mean and standard error
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight), se = sd(Weight)/sqrt(n) ) %>%
  ungroup()

# Looking at the data:
glimpse(RATSS)



# Plotting the mean profiles:
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.9,0.3)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
```



```{r}
# Creating summary data
RATSL_2 <- RATSL %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()

# Looking at the data
glimpse(RATSL_2)

```
```{r}
# Drawing boxplots
ggplot(RATSL_2, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "blue") +
  scale_y_continuous(name = "mean(Weight)")
```

We see the outliers of each group in the plot.There are three outliers. For the first and third group they are the smallest values and for the second group - the biggest value.


```{r}
RATSL_2 %>%
  group_by(Group) %>%
  summarize(max(mean))
```


```{r}
RATSL_2 %>%
  group_by(Group) %>%
  summarize(min(mean))
```

So for the first group the outlier value is 238.9, for the third group: 495.2 and for the second group: 594.

```{r}
# Creating a new data by filtering the outliers
RATSL_3 <- RATSL_2 %>%
  filter(mean !=238.9 & mean != 495.2 & mean != 594)
```


```{r}
# Drawing a boxplot
ggplot(RATSL_3, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "blue") +
  scale_y_continuous(name = "mean(Weight)")
```

Now the data is without outliers.

Let's perform a t-test to see if group 1 and group 2 differs.

```{r}

# Filtering data to have group 1 and group 2
RATSL_2groups <- filter(RATSL_3,(Group==1 | Group==2))

# Performing a two-sample t-test
t.test(mean ~ Group, data = RATSL_2groups, var.equal = TRUE)


```

Based on the result of the t-test, we see that at 95% confidence level, there is a significant difference between two means. As the p value is lower that 0.05 the null hypothesis (that the means are equal) is rejected.


```{r}
# Adding the baseline from the original data as a new variable to the summary data

RATS <- read.csv('data/RATS.csv')

RATSL_4 <- RATSL_2 %>%
  mutate(baseline = RATS$WD1)

# Fitting the linear model
fit <- lm(mean ~ baseline + Group, data = RATSL_4)

# Computing the analysis of variance table for the fitted model with anova()
anova(fit)

```



## Part 2. BPRS data



```{r}
# Reading the data set
BPRSL <- read.csv('data/BPRSL.csv')

# Looking at the structure of the data set
str(BPRSL)

# Converting variables as factors
BPRSL$treatment <- factor(BPRSL$treatment)
BPRSL$subject <- factor(BPRSL$subject)


```







