# Dimensionality reduction techniques


```{r}
date()
```

## Data

In this chapter the *human* dataset, which originates from the [United Nations Development Programme](http://hdr.undp.org/en/content/human-development-index-hdi), is analyzed. 
The code of data preparation along with the data is in my [Github repository](https://github.com/ritabei/IODS-project/blob/master/data/create_human.R).


Let's load and explore the data.

```{r}
# Loading the data
human <- read.table('data/human.csv',sep = ',',header = T, row.names = 1)

# Looking at the structure of the dataset
str(human)
dim (human)
names(human)
```

The data set consists of 155 observations and 8 variables. Each observation has a name which indicates a country. The variables are:

*Edu2.FM*: ratio of proportion of females with at least secondary education and proportion of males with at least secondary education.

*Labo2.FM*: ratio of proportion of females in the labour force and proportion of males in the labour force.

*Life.Exp*: life expectancy at birth.

*Edu.Exp*: expected years of schooling.

*GNI*: gross National Income per capita.

*Mat.Mor*: maternal mortality ratio.

*Ado.Birth*: adolescent birth rate.

*Parli.F*: percetange of female representatives in parliament.


```{r}
# Looking at the summary of the variables
summary(human)
```
All of the variables have different scales. The features, especially, *GNI*, *Mat.Mor*, *Ado.Birth* exhibit high variability.


Let's look at the pairwise relationships between the variables.
```{r}
# Plotting a matrix of the variables
pairs(human)

```
```{r}
library(corrplot)
library(dplyr)
# Computing the correlation matrix
cor(human) %>% corrplot(type = "upper", tl.cex=0.7)

```


Many variables seem to be correlated. For example, when *Life.Exp* (life expectancy at birth) increases, *Edu.Exp* (expected years of schooling) and *GNI* increase as well. With increasing *Life.Exp*, *Mat.Mor* (maternal mortality ratio) and *Ado.Birth* decreases (adolescent birth rate).


Let's look at the distribution of the variables.

```{r, fig.height = 4, fig.width = 4 }
library(ggplot2)
# Ploting a histogram of life expectancy at birth variable
ggplot(human, aes(x=Life.Exp)) +theme_minimal() + geom_histogram(color="darkblue", fill="lightblue", binwidth=5) + ggtitle("Life expectancy at birth")
```


```{r, fig.height = 4, fig.width = 4 }
library(ggplot2)
# Ploting a histogram of life expectancy at birth variable
ggplot(human, aes(x=Edu.Exp)) +theme_minimal() + geom_histogram(color="darkblue", fill="lightblue", binwidth=1) + ggtitle("Expected years of schooling")+ stat_function(fun = function(x) dnorm(x, mean = mean(human$Edu.Exp), sd = sd(human$Edu.Exp)) * 1 * nrow(human))
```

```{r, fig.height = 4, fig.width = 4 }
library(ggplot2)
# Ploting a histogram of Mat.Mor variable
ggplot(human, aes(x=Mat.Mor)) +theme_minimal() + geom_histogram(color="darkblue", fill="lightblue", binwidth=50) + ggtitle("Maternal mortality ratio")
```

```{r, fig.height = 4, fig.width = 4 }
library(ggplot2)
# Ploting a histogram of Ado.Birth variable
ggplot(human, aes(x=Ado.Birth)) +theme_minimal() + geom_histogram(color="darkblue", fill="lightblue", binwidth=10) + ggtitle("Adolescent birth ratio")
```


```{r, fig.height = 4, fig.width = 5 }
library(ggplot2)
# Ploting a histogram of Parli.F variable
ggplot(human, aes(x=Parli.F)) +theme_minimal() + geom_histogram(color="darkblue", fill="lightblue", binwidth=5) + ggtitle("Percetange of female representatives in parliament")
```

Out of all distributions, only expected years of schooling resembles a normal distribution. Maternal mortality ratio is mostly quite low. However, there are very high ratio exceptions. Life expectancy has most counts between 70 and 80 years. In very few cases females take up more than 50% of parliament places.


## Principal Component Analysis (PCA)

In this section PCA is applied on non-standartized and standartized and results are compared at the end of the section.

### Non-standartized data


```{r message=FALSE}
options(warn = -1)
# Performing PCA (with the SVD method)
pca_human_ns <- prcomp(human)

# Creating and printing out a summary of pca_human
s <- summary(pca_human_ns)
s

# Rounded percentages of variance captured by each PC
pca_pr <- round(100*s$importance[2,], digits = 1) 

# Printing out the percentages of variance
pca_pr

# Creating an object pc_lab to be used as axis labels
pc_lab <- paste0(names(pca_pr), " (", pca_pr, "%)")

# Drawing a biplot
biplot(pca_human_ns, cex = c(0.8, 1), col = c("grey40", "deeppink2"), xlab = pc_lab[1], ylab = pc_lab[2], sub = 'Gross National Income per capita (GNI) has the highest variability out of all features and it highly contributes to PC1 component.', cex.sub = 0.7)







```

### Standartized data

```{r message=FALSE}
options(warn = -1)
# Standardizing the variables
human_std <- scale(human)


# Performing PCA (with the SVD method)
pca_human <- prcomp(human_std)


# Creating and printing out a summary of pca_human
s <- summary(pca_human)
s

# Rounded percentages of variance captured by each PC
pca_pr <- round(100*s$importance[2,], digits = 1) 

# Printing out the percentages of variance
pca_pr

# Creating an object pc_lab to be used as axis labels
pc_lab <- paste0(names(pca_pr), " (", pca_pr, "%)")

# Drawing a biplot
biplot(pca_human, cex = c(0.8, 1), col = c("grey40", "deeppink2"), xlab = pc_lab[1], ylab = pc_lab[2],sub= 'Standartized variables more equally contribute to the principal components.', cex.sub=0.7)


```

Standardization of the data set makes a very big difference. Without standardizing, very high variability in *GNI* highly contributes to the first principal component and thus, all variability in the data is explained only with PC1.
When data is standardized, all variables contribute more equally to the components and the first principal component explains over 50% of variability in the data.

The angles between features in the biplot of standardized data are small. They represent high correlation between those features.

Ratio of proportion of females in the labour force and proportion of males in the labour force as well as percetange of female representatives in parliament are the only two variables which contribute to the second principal component. These two variables help to explain less variance in the data compared to education, life expectancy, maternal mortality and adolescent birth ratio variables.



## Multiple Correspondence Analysis

In this section the *tea* dataset from the package *FactoMineR* is used. It contains the answers of a questionnaire on tea consumption.

```{r}
# Loading tea dataset
library(FactoMineR)
data('tea')

# Looking at the structure of the dataset
str(tea)
dim(tea)

# Looking at the summaries of the variables
summary(tea)
```

This dataset has 300 observations and 36 variables. As there are many variables, I will selecte only some of them for further analysis.

age_Q, sex, price, sugar, frequency, 


```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
# Subsetting the dataset
selected_columns <- c ('age_Q', 'sex', 'price', 'sugar', 'frequency', 'relaxing' )
tea_selected <- dplyr::select(tea, one_of(selected_columns))

# Visualize the selected dataset
gather(tea_selected) %>% ggplot(aes(value)) + facet_wrap("key", scales = "free") + geom_bar() + theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))




```

It seems that the highest tea consumption is in the age group of 15-24 year. The cheap tea isn't chosen often. More people find tea drinking relaxing and more female than male drinks tea. Tea with no sugar is slightly more preferred.


Let's do the multiple correspondence analysis.

```{r}
# Multiple correspondence analysis
mca <- MCA(tea_selected, graph = FALSE)

# Summarizing the model
summary(mca)

# Visualizing the MCA
plot(mca, invisible=c("ind"), habillage = "quali")
```

The horizontal dimension explains 10.79% of the variance in the data whereas the vertical dimension explains only 9.73%. Together both dimension explain only around 20% of variance in the data set.
 
The graph in MCA presents the information in a condense form where the distance between variable categories gives a measure of their similarity.


Let's look at some possible conclusions and see if they can be drawn from the data as well.

1) Tea with no sugar is more chosen by people from two older age groups.

```{r}

# Initializing a plot of alc_use and going out
ggplot(tea_selected, 
       aes(x = age_Q, 
           fill = sugar )) + 
  geom_bar(position = position_dodge(preserve = "single")) +labs(
   title = 'Sugar preference in different age group') + theme_minimal()
```
Yes, from the barplot we can see, that more peaople in youger age groups prefer tea with sugar and more people in older groups, prefer the tea without sugar.


2) Female are more likely to drink tea twice a day.



```{r}
# Initializing a plot of alc_use and going out
ggplot(tea_selected, 
       aes(x = sex, 
           fill = frequency )) + 
  geom_bar(position = position_dodge(preserve = "single")) +labs(
   title = 'Tea drinking frequencies famale vs male') + theme_minimal()
```

Seems that this conclusion is visible in the data as well.



3) Sugar is chosen by people who drink tea least often (1 to 2 times per week).

```{r}
# Initializing a plot of alc_use and going out
ggplot(tea_selected, 
       aes(x = frequency, 
           fill = sugar )) + 
  geom_bar(position = position_dodge(preserve = "single")) +labs(
   title = 'Sugar preference in different age group') + theme_minimal()
```

It seems that this conclusion can be drawn from the data as well.

